Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-04-02 analysis/paper/paper.Rmd
minor edit and grammar check

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that both millet and rice have been introduced into southwest China at the latest in 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic period. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on the site at Guijiabao located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). The Guijiabao site allows us to understand the early adoption of agriculture because of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of the Guijiabao site can be divided into three temporal phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a]. Plant remains reveal the cultivation of crops dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022]. This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022]. In addition to millet, a smaller number of rice grains (*Oryza sativa*) have also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences [@Huan2022].

<!--info about Gaoshan-->
To compare culinary practices for different ecological zones, we examined a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to similar pottery forms and styles, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Gaoshan2017]. The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020]. It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from the Middle and Lower Yangzi valleys [@Zhang2010].

Although the recent findings at Guijiabao and Gaoshan support mixed agriculture, it is still unclear how much millet or rice was included in local culinary practices during the Neolithic period. The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwest China reveal a predominance of rice, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017] and Henglanshan and Baiyangcun in Yunnan [@Jiang2016; @Henglanshan2014]. This indicates there were variations in agricultural practices in southwest China. To better understand the extent of the adoption of new crops, organic residue analysis for ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

## Samples
We selected ceramic fragments that were identified as typical types from Guijiabao and Gaoshan. The fragments were from known contexts with radiocarbon dates: Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)). Ceramic fragments from Guijiabao were recovered from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season. The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations. Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Guijiabao2017a; @Hao2022]. This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.

@@

knitr::kable(Ch_sherds,
             caption = "A list of the pottery sherds that were analyzed in this study. The sherd labeled 'SYG' means potsherds from Guijiabao, while the label 'CDG' refers to Gaoshan")
```

## Methods
We detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS) and quantified lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

For the sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C). We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS.

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol, respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed by GC-C-IRMS. With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might from solvent and reagent. Internal standards were added before and after the lipid extraction for quantification.

The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector. Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness). Helium carrier is at a constant flow rate of 1.3 ml/min. The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes. The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode. The carrier gas was helium at 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was held at 90°C for 1.3 minute, and then raised to 320°C at a rate of 5°C/min. The final temperature was held for 22 minutes. The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface. The sample injection volume was 1 ul in toluene and was injected in splitless mode. Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set at 90 °C for 1.5 minutes and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold. Helium was used as a gas carrier in constant flow mode at 1.24 ml/minute.

@@

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but in relatively small amounts in the majority of them (Table \@ref(tab:FA-composition-table)]. The lipid profiles show that the main compounds in most potsherds are saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from larger molecules, such as triacylglycerols, after long-term burial processes due to oxidation and microbial degradation [@Dudd1998; @Eerkens2007]. High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005]. However, we may still need to examine their isotopic values to confirm their sources.

@@

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao and Gaoshan. Cn:x refers to fatty acid with n carbon atoms and x double bonds. IS indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of ceramic or a different use. This sample also has a charred interior surface, indicative of a function associated with food processing. Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the Guijiabao potsherd shows the presence of long-chain fatty acids. Cn:x refers to fatty acid with n carbon atoms and x double bonds. IS indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
```

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

The results show that the majority of Guijiabao potsherd samples contain fatty acids from ruminant adipose fats, with one potsherd (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaoshan (CDG-062) demonstrates ruminant adipose fats as well. The potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low 13C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting those ruminant animals seemed to have consumed mainly C3 plants. There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value, suggesting a C4-based diet for the ruminant animal.

```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C16:0 and C18:0. B: The Δ13C (δ13C18:0-16:0) values plotted against δ13C16:0 values"}
knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
```

>>>>

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or terrestial animals across areas in southwest China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in those areas. It is widely accepted that both millet and rice were introduced into southwest China in 4700 BP at the latest. However, to what extent new crops were used by local peoples at an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis. The results show signals of terrestrial animals in most potsherds and a C3 plant source in one potshard, during the Neolithic period. Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming. This study also demonstrates an approach using organic residue analysis to explore pottery use and culinary practices during the Neolithic period in southwest China. A more systematic sampling of different types of potsherds and temporal phases with larger samples is needed to provide a clearer picture of ancient diets and subsistence.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on the Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). The Guijiabao site allows us to understand the early adoption of agriculturebecause of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao was a settlement site [@Hao2022]. The chronology of the Guijiabao site can be divided into three phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles that were identified [@Hao2022].

<!--the major finding at GJB-->
The common vessel forms unearthed at the Guijiabao site include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between the Guijiabao site and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a]. Plant remains reveal that cultivated crops were dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022]. This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022]. In addition to millet, a smaller number of rice grains (*Oryza sativa*) has also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences [@Huan2022].

<!--info about Gaoshan-->
To compare the culinary practices in different ecological zones, we examined a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at the Gaoshan site reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and archaeological features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to similar pottery forms and styles, the Gaoshan site can be dated to the early Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Gaoshan2017]. The primary subsistence strategy at the Gaoshan site could have been rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020]. It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from people moving from the Middle and Lower Yangzi valleys to the Sichuan area [@Zhang2010].

Although the recent findings at the Guijiabao and Gaoshan sites support mixed agriculture as the primary subsistence strategy, it is still unclear how much millet or rice was included in the local diets during the Neolithic period. The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwest China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and at Henglanshan and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014]. This indicates that there were variations in agricultural practices in southwest China. To better understand the extent of the adoption of new crops, organic residue analysis of ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

## Samples
We selected ceramic fragments that were identified as typical from the Guijiabao and Gaoshan sites. The fragments date to the Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)), which were determined from known contexts based on radiocarbon dating. Ceramic fragments from the Guijiabao site were excavated from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season. The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations. Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that these ceramics were used in daily life [@Guijiabao2017a; @Hao2022]. This research also attempts to confirm that these pottery vessels were used as cooking/storage wares for food.

@@

knitr::kable(Ch_sherds,
             caption = "A list of the pottery sherds that were analyzed in this study. The sherd labeled 'SYG' indicates potsherds from the Guijiabao site, while the label 'CDG' refers to those from the Gaoshan site")
```

## Methods
We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), which are found in most animals and plants, we identified their stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

For sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C). We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS.

The remaining 95% aliquot was fractioned using solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed using GC-C-IRMS. With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might have come from solvents and reagents. Internal standards were added before and after the lipid extraction for quantification.

The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector. Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness). The carrier gas was helium set at a constant flow rate of 1.3 ml/min. The temperature of the oven was set at 90 °C and held for 1 minute and then was increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes. The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode. The carrier gas was helium flowing at a rate of 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was set at 90°C and held for 1.3 minute, and then was raised to 320°C at a rate of 5°C/min. The final temperature was set for 22 minutes. The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface. The sample injection volume was 1 ul in toluene and was injected in splitless mode. Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set at 90 °C for 1.5 minutes and then was raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold. Helium was used as a gas carrier with a constant flow rate of 1.24 ml/minute.

@@

The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the presence of lipid compounds were above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)]. The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)). These fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from larger molecules, such as triacylglycerols, due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-odd carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds had higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005]. However, we may still need to examine their isotopic values to confirm their sources.

@@

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from the Guijiabao and Gaoshan sites. "Cn:x" refers to fatty acid with n carbon atoms and x double bonds. "IS" indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the samples. A high yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation. This sample also has a charred interior surface, indicative of a function associated with food processing. Long-chain fatty acids from C20:0 to C32:0 are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high amount of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the potsherd from the Guijiabao site shows the presence of long-chain fatty acids. "Cn:x" refers to fatty acid with n carbon atoms and x double bonds. "IS" indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
```

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 have a distribution between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

The results show that the majority of the potsherd samples from the Guijiabao site contain fatty acids from ruminant adipose fats, with one potsherd sample (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A). Analysis of the potsherd from the Gaoshan site (CDG-062) indicates ruminant adipose fats as well. The potsherds (SYG-TN13-E22-1 and SYG-3) from the Guijiabao site dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of these two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low 13C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for determining food categories.

In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting these ruminant animals seemed to have consumed mainly C3 plants. There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively large δ^13^C value, suggesting a C4-based diet for the ruminant animal.

```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C16:0 and C18:0. B: The Δ13C (δ13C18:0-16:0) values are plotted against δ13C16:0 values"}
knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
```

# Discussion
<!--A brief summary of interesting results-->
The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets) or C3 plants (e.g., rice), or animals (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period. Our results show that food residues in potsherds from the sites at Guijiabao and Gaoshan were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to discover that non-ruminant fats in a Guijiabao potsherd might have originated from C3 plants (see below). Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that these ceramics were used as cooking or storage vesssels. The low lipid yield could be attributed to the bottom of the vessels, where there were fewer residues than other parts of the vessel, for example, the rim [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found that one sample from the vessel bottom dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples, reflecting better preservation or a much higher lipid yield. We recognize that there are limitations and challenges for determining lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. In addition, a detailed analysis of faunal remains from the Guijiabao site is still in progress, which may constrain a comprehensive comparison between different types of evidence. Nevertheless, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices at the Guijiabao site.
